---
title: "Sentence-case title naming the visual lesson, not just the dataset"
subtitle: "Session N · TidyTuesday YYYY-MM-DD · Grammar layer: <layer>"
date: <session date>
---

<!--
  TEMPLATE — copy to a new .qmd, do not render this file (leading underscore keeps
  Quarto from rendering it). Every section below is part of the genre: a record of
  decisions, not an essay. Delete these comments as you fill it in.
-->

::: {.callout-tip icon=false}
## Session N · YYYY-MM-DD · facilitated by [facilitator(s)]
With [named consenting participants — only names present in
`sessions/YYYY-MM-DD/consent.md`]. Design decisions were made live by the group; the
writeup was drafted by [actual agent and model] from [specs and decision notes;
recording/transcript only if used] and approved before publishing.
[How these sessions work](index.qmd).
:::

One or two opening paragraphs: what the dataset is, and the question the group settled on.

::: {.callout-tip collapse="true"}
## Reproducing this page
Everything runs from CSVs committed in this repository. To refresh from source:

```r
# download.file(...) lines for the raw pull
```

Packages: `tidyverse`. Source: [name](link), via TidyTuesday
[YYYY-MM-DD](https://github.com/rfordatascience/tidytuesday/tree/main/data/YYYY/YYYY-MM-DD).
:::

```{r setup}
library(tidyverse)
theme_set(theme_minimal(base_size = 13))
# read from data/<slug>.csv
```

## The grammar layer

A short passage connecting this session to its layer of the grammar of graphics: what the
layer does, and why this dataset stresses it.

## Design decisions

The heart of the post. For each major decision, record it **as a decision**: what was
considered, what was chosen, and why. Quote the gg-spec YAML as it stood at each step —
the spec is the record.

```yaml
# spec as it stood after the first round of discussion
```

```{r fig-first}
# the figure that spec compiles to
```

Then the revision: what the group saw in the rendered figure that changed the spec, and
the next spec/figure pair. Two or three such rounds is typical.

## Pain points

What fought back: the reshaping that took three tries, the legend that wouldn't behave,
the scale that misled. Include what *looking at the render* caught that the code review
didn't.

## Lessons learned

Three to five short, portable statements a reader could apply to their own next chart.

## What the data can't say

The limits passage: the confounder, the censoring, the too-short record. Every page earns
one.
