---
title: "One mapping line, two opposite conclusions"
subtitle: "A worked example of the gg-spec round trip · Palmer penguins, TidyTuesday 2020-07-28"
date: 2026-08-18
---

::: {.callout-tip icon="false"}
## Not a session page — a demonstration

This page shows the working method the sessions use: a plot is specified as a short YAML document of grammar-of-graphics mapping rules, the spec is compiled to ggplot2 code, the figure is rendered and *looked at*, and what looking reveals drives the next revision of the spec. [How these sessions work](index.qmd).
:::

In a live session, nobody types ggplot2 code to begin with. The group dictates **mapping rules** — which variable goes to which aesthetic, which geoms draw them, which scales transform them — and those rules are written down as a spec. The spec is short enough to argue about on a shared screen, and it is the record of the design decisions that ends up quoted in the writeup.

Here is one full round trip on the Palmer penguins data.

::: {.callout-tip collapse="true"}
## Reproducing this page

Runs from `data/penguins.csv`, committed in this repository. To refresh from source:

``` r
download.file(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2020/2020-07-28/penguins.csv",
  "data/penguins.csv")
```

Packages: `tidyverse`. Source: Horst, Hill & Gorman, [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/), via TidyTuesday [2020-07-28](https://github.com/rfordatascience/tidytuesday/tree/main/data/2020/2020-07-28).
:::

```{r setup}
library(tidyverse)
theme_set(theme_minimal(base_size = 13))

penguins <- read_csv("data/penguins.csv", show_col_types = FALSE) |>
  filter(!is.na(bill_depth_mm))

slope_pooled <- coef(lm(bill_depth_mm ~ bill_length_mm, data = penguins))[2]
slopes_species <- penguins |>
  group_by(species) |>
  group_modify(~ tibble(slope = coef(lm(bill_depth_mm ~ bill_length_mm, data = .x))[2])) |>
  ungroup()
```

## Round 1: the spec as first dictated

The group asks a simple question — do penguins with longer bills also have deeper bills? — and dictates the minimal spec: two positional mappings, points, and a linear trend.

```{r spec-1, echo=FALSE, results='asis'}
cat("```yaml\n", paste(readLines("specs/penguins-example.yml"), collapse = "\n"), "\n```", sep = "")
```

The spec compiles to code a reader could have written by hand — layers in spec order, nothing the spec didn't ask for:

```{r fig-round-1}
#| fig-cap: "Round 1: bill depth against bill length, all penguins pooled. The fitted line slopes down."
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title    = "Do longer bills mean deeper bills?",
    subtitle = sprintf("%d penguins, Palmer Archipelago 2007-09", nrow(penguins)),
    x = "Bill length (mm)", y = "Bill depth (mm)"
  )
```

## Looking at the render

The fitted line says *no*: pooled across all `r nrow(penguins)` birds, each extra millimetre of bill length goes with `r sprintf("%.3f", abs(slope_pooled))` mm *less* depth. That contradicts any intuition about bigger birds having bigger bills — and looking at the point cloud shows why the line is suspect: the points sit in distinct clumps the spec has done nothing to distinguish.

The group's diagnosis, in grammar terms: **a variable that structures the data has been left out of the mapping**. The fix is one line.

## Round 2: one mapping line added

```{r spec-2, echo=FALSE, results='asis'}
cat("```yaml\n", paste(readLines("specs/penguins-example-2.yml"), collapse = "\n"), "\n```", sep = "")
```

Only `colour: species` (and a colour-blind-safe scale for it) has been added. But because `geom_smooth` inherits the plot-level mapping, colour doesn't just tint the points — it **groups the stat**, and one regression becomes three.

```{r fig-round-2}
#| fig-cap: "Round 2: the same data with species mapped to colour. Within every species the slope is positive."
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, colour = species)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_colour_manual(values = unname(palette.colors(3, palette = "Okabe-Ito"))) +
  labs(
    title    = "Within every species, the slope reverses",
    subtitle = sprintf("The same %d penguins, now mapped by species", nrow(penguins)),
    x = "Bill length (mm)", y = "Bill depth (mm)", colour = "Species"
  )
```

Within each species the relationship runs the way intuition says it should — slopes of `r paste(sprintf("%.2f", slopes_species$slope), collapse = ", ")` mm per mm for `r paste(slopes_species$species, collapse = ", ")` respectively, every one positive. The pooled negative slope was Simpson's paradox: Gentoos are long-billed but shallow-billed, and pooling let the between-species contrast masquerade as a within-penguin trend.

## What the round trip teaches

- **Mappings are decisions.** The difference between the two figures — and two opposite conclusions — is a single line in the spec. Leaving a variable out of the mapping is as much a choice as putting it in.
- **Aesthetics reach further than appearance.** Mapping colour regrouped the statistical layer, not just the ink. In the grammar, mappings feed *stats*, not only geoms.
- **The render is part of the method.** The clumpy point cloud was visible in Round 1's figure but nowhere in its spec or code. You find it by looking.

## What the data can't say

Three species from three islands over three seasons cannot separate species from island effects (Chinstraps and Gentoos were each sampled on effectively one island), and a within-species slope across individuals is still not a growth trajectory — no penguin was measured twice. The paradox here is a lesson about mappings, not a finding about penguins.